Classless Java

### 2016-03-24 OOPSLA16/predef/java.tex
Section 6

\lstdefinelanguage{JavaScala}{
  morekeywords={public,int,interface,implements,default,
    abstract,case,void,catch,class,def,static,%
    do,else,extends,false,final,finally,%
    for,if,implicit,import,match,mixin,%
    new,null,object,override,package,%
    private,protected,requires,return,sealed,%
    super,this,throw,trait,true,try,%
    type,var,while,yield},
  otherkeywords={=>,<-,<\%,<:,>:,\#,@},
  sensitive=true,
  morecomment=[l]{//},
  morecomment=[n]{/*}{*/},
  morestring=[b]",
  morestring=[b]',
  morestring=[b]"""
}

>>>>

\lstdefinelanguage{JavaScala}{
  morekeywords={public,int,interface,implements,default,
    abstract,case,void,catch,class,def,static,%
    do,else,extends,false,final,finally,%
    for,if,implicit,import,match,mixin,%
    new,null,object,override,package,%
    private,protected,requires,return,sealed,%
    super,this,throw,trait,true,try,%
    type,var,while,yield,obj,updater},
  otherkeywords={=>,<-,<\%,<:,>:,\#,@},
  sensitive=true,
  morecomment=[l]{//},
  morecomment=[n]{/*}{*/},
  morestring=[b]",
  morestring=[b]',
  morestring=[b]"""
}

### 2016-03-24 OOPSLA16/sections/FutureWork.tex
Section 6

\section{Future work}\label{sec:futurework}
\begin{comment}
\subsection{Encapsulation}
The biggest limitation of our approach is the inability to model visibility restrictions. For example, the absence
of support for private/protected methods in Java 8 interfaces forces all
members of interfaces to be public, including static methods. Since we use
abstract methods to encode state, our state is always all public. Still, because
the state can only be accessed by methods, it
is impossible for the user to know if a certain method maps directly to a field
or if it has a default implementation.  If the user wants a constructor that
does not directly maps to the fields, (as for secondary constructors in Scala)
he can simply define its own \Q@of@ method and delegate on the generated one:
\begin{lstlisting}
@Obj interface Point {  int x(); int y();
    static Point of(int val){return Point.of(val,val);} }
\end{lstlisting}
However, the generated \Q@of@ method would also be present and public.  If a
future version of Java was to support \emph{static private methods in
  interfaces} we could extend our code generation to handle also encapsulation.

@@

As shown in~\cite{BettiniDSS13} interfaces are enough to support encapsulation/privateness by a
 variation of the facade pattern:
\begin{lstlisting}
@Obj interface InternalData implements ExposedData {
    /*all the methods*/} //package visibility
public interface ExposedData {
  /*just the exposed interface*/
  static ExposedData of(...) {
    return InternalData.of(...);}  }
\end{lstlisting}

This is a safe and explicit way to list all the public interfaces and to hide all the implementation
details from outside of the package.\footnote{
Note how \Q@ExposedData.of@ do not need to map directly to the fields, and can also perform
invariant checking.}

Using packages as a boundary we can make all object interfaces \emph{where some form of privateness is desirable}
as invisible types for the rest of the program.
This preserves encapsulation, but limits inheritance:
package protected interfaces cannot be seen,
thus cannot be implemented from outside the package.

This can raise questions: keeping fields private is a staple of CB \objectoriented programming;
by forcing the user/heir to rely only on methods, we avoid (a part)
of the fragile base class problem and we obtain representation independence.

In IB fields do not exists; abstract state methods like getters and setters are abstract:
it does not make sense to make them private; however, since they are methods,
the base interface may be able to evolve freely.
To understand if a fragile base interface problem exists in IB, we need more research.

@@

%\subsection{Clone, toString, equals and hashCode}
\paragraph{Clone, toString, equals and hashCode}
Methods originally defined in Java class \Q@Object@, such as \Q@clone@ and
\Q@toString@, can be supported by our approach with special care. If an
interface annotated with \mixin asks an implementation for \Q@clone@,
\Q@toString@, \Q@equals@ or \Q@hashCode@ we can easily generate one from the
fields.\footnote{In particular, for clone we can do automatic return type
  refinement as we do for \Q@with-@ and fluent setters. Note how this would
  solve most of the Java ugliness related to \Q@clone@ methods.}  However, if
the user wishes to provide his own implementation, since the method is also
implemented in \Q@Object@, a conflict arises. The generated code can resolve the
conflict inside \Q@of@, by implementing the method and delegating it to the user
implementation, thus

@@

\begin{lstlisting}
interface Point { ...//as before
    public static Point of(int _x, int _y) {
        return new Point() {...
            public Point clone() {
                return Point.super.clone();}};} }
\end{lstlisting}

\paragraph{A real IB language}
In this article we presented the concept of IB as a programming pattern over Java.
An alternative would be to have a language providing IB as its only paradigm.
We do not wish to define here a complete IB language, just to show the reader IB as
a promising way to design a new OO language.
To define such language, the first task is to define the set of abstract state operations.
Of course we need getters; \Q@withX@ methods seams to be pretty convenient, and they enable a
clean function programming style that would be otherwise too cumbersome.

Currently we are offering both setters and fluent setters.
We believe be fluent setters to be sufficient; we could drop the concept of (non fluent)setters.

Manually defining abstract state methods can be verbose, we could offer a convenient syntax to
declare in a compact way get/set/with.
For example
\begin{lstlisting}
obj interface Foo{
  String bar;// exapanded as getter and with:
  //String bar(); Foo withBar(String val);
  var String beer;//as before plus fluent setter
  }
\end{lstlisting}

It could be possible to offer pre initialized abstract state, for example
\begin{lstlisting}
obj interface Foo{
  var String bar="hello";
  }
\end{lstlisting}

could be a short hand for writing

@@

We may also want to offer a compact syntax for property updaters and generalized with methods;
\begin{lstlisting}
obj interface Foo{
  updater Bar, Beer//declares imperative updaters:
  //Foo set(Bar val); Foo set(Beer val);
  with Foo, Beer//declared functional updaters:
  //Foo with(Bar val); Foo with(Beer val);
  }
\end{lstlisting}
Finally, certain methods may have a specific meaning, for example
if a method called \Q@postInit@ is present, it will be called after construction,
and it may be handled specially during multiple implementations, so that all relevant \Q@postInit@ methods are
called in some order.

>>>>

\section{Future work}\label{sec:futurework}

This section discusses some possible future work, including native
language support for IB.

\begin{comment}

\subsection{Encapsulation}

The biggest limitation of our approach is the inability to model visibility restrictions. For example, the absence
of support for private/protected methods in Java 8 interfaces forces all
members of interfaces to be public, including static methods. Since we use
abstract methods to encode state, our state is always all public. Still, because
the state can only be accessed by methods, it
is impossible for the user to know if a certain method maps directly to a field
or if it has a default implementation.  If the user wants a constructor that
does not directly maps to the fields, (as for secondary constructors in Scala)
he can simply define its own \Q@of@ method and delegate on the generated one:
\begin{lstlisting}
@Obj interface Point {  int x(); int y();
    static Point of(int val){return Point.of(val,val);} }
\end{lstlisting}
However, the generated \Q@of@ method would also be present and public.  If a
future version of Java was to support \emph{static private methods in
  interfaces} we could extend our code generation to handle also encapsulation.

@@

A limitation of our approach, inherited from Java, is the lack of some
support for encapsulation: in Java interfaces all methods are public.
However, it is possible to workaround the
limitations of interfaces by a
 variation of the facade pattern~\cite{BettiniDSS13}:
\begin{lstlisting}
@Obj interface InternalData implements ExposedData {
    /*all the methods*/} //package visibility
public interface ExposedData {
  /*just the exposed interface*/
  static ExposedData of(...) {
    return InternalData.of(...);}  }
\end{lstlisting}

\noindent This is a safe and explicit way to list all the public interfaces and to hide all the implementation
details from outside of the package.\footnote{
Note how \Q@ExposedData.of@ does not need to map directly to the fields, and can also perform
invariant checking.}
Using packages as a boundary we can make all object interfaces \emph{where some form of privateness is desirable}
as invisible types for the rest of the program.
This preserves encapsulation, but limits inheritance:
package protected interfaces cannot be seen,
thus cannot be implemented from outside the package.
This can raise questions: keeping fields private is a staple of CB.
By forcing the user/heir to rely only on methods, we avoid (a part)
of the fragile base class problem and we obtain representation independence.
In IB fields do not exists; abstract state methods like getters and
setters are abstract: it does not make sense to make them private;
however, since they are methods, the base interface may be able to
evolve freely.  To understand if a fragile base interface problem
exists in IB, we need more research.

@@

%\subsection{Clone, toString, equals and hashCode}

\begin{comment}
\paragraph{Clone, toString, equals and hashCode}
Methods originally defined in Java class \Q@Object@, such as \Q@clone@ and
\Q@toString@, can be supported by our approach with special care. If an
interface annotated with \mixin asks an implementation for \Q@clone@,
\Q@toString@, \Q@equals@ or \Q@hashCode@ we can easily generate one from the
fields.\footnote{In particular, for clone we can do automatic return type
  refinement as we do for \Q@with-@ and fluent setters. Note how this would
  solve most of the Java ugliness related to \Q@clone@ methods.}  However, if
the user wishes to provide his own implementation, since the method is also
implemented in \Q@Object@, a conflict arises. The generated code can resolve the
conflict inside \Q@of@, by implementing the method and delegating it to the user
implementation, thus

@@

\begin{lstlisting}
interface Point { ...//as before
    public static Point of(int _x, int _y) {
        return new Point() {...
            public Point clone() {
                return Point.super.clone();}};} }
\end{lstlisting}
\end{comment}

\paragraph{A real IB language}
In this paper we presented the concept of IB as a programming pattern over Java.
An alternative would be to have a language providing IB as its only paradigm.
We do not wish to define here a complete IB language, just to show the reader IB as
a promising way to design a new OO language.
To define such language, the first task is to define the set of abstract state operations.
Of course we need getters; \Q@withX@ methods are also convenient, and they enable a
clean functional programming style that would be otherwise too cumbersome.
Currently we are offering both setters and fluent setters.
We believe fluent setters to be sufficient; we could drop the concept of (non fluent)setters.
Manually defining abstract state methods can be verbose, we could offer a convenient syntax to
declare in a compact way get/set/with.
For example:
\begin{lstlisting}
obj interface Foo{
  String bar;// expanded as getter and with:
  //String bar(); Foo withBar(String val);
  var String beer;//as before plus fluent setter
}
\end{lstlisting}

\noindent It should be possible to offer pre initialized abstract state, for example
\begin{lstlisting}
obj interface Foo{ var String bar="hello"; }
\end{lstlisting}
\noindent could be a short hand for writing

@@

We may also want to offer a compact syntax for property updaters and generalized with methods;
\begin{lstlisting}
obj interface Foo{
  updater Bar, Beer;//declares imperative updaters:
  //Foo set(Bar val); Foo set(Beer val);
  with Foo, Beer;//declared functional updaters:
  //Foo with(Bar val); Foo with(Beer val);
}
\end{lstlisting}
Finally, certain methods may have a specific meaning. For example
if a method called \Q@postInit@ is present, it will be called after construction.
It may be handled specially during multiple implementations, so that all relevant \Q@postInit@ methods are
called in some order.
